---
title: "World Happiness Report from year 2005-2021"
output: slidy_presentation
author: "Kristen Blessman-Garcia"
date: "2023-04-27"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = F, message = F)
data1 <- read.csv("C:\\Users\\diego\\Desktop\\data science folder\\World Happiness Report 2005-2021.csv", header=TRUE, stringsAsFactors=FALSE)
```


 
## Introduction
The data is a CSV from kaggle based on the Gallup Poll/World Happiness Reports from 2005-2021. 

## Background
I have practiced as an occupational therapy assistant for the past ten years and have an interest in the social/emotional well being of my clients.  My domain focus for data science is psychology.

---

## Questions 

- To understand global happiness trends and the factors contributing to well-being.  

  - Which factors are related to an increased score on the happiness report?
  - Which countries were in the top ten most frequently?
  - Which countries had the lowest happiness score most frequently?

---

## World Happiness Report measured in Life.Ladder 

Please imagine a ladder, with steps numbered from 10 at the top to 0 at the bottom.

## Ten

  - The top of the ladder represents the best possible life for you. 
  -
  -
  -
  -
  -

##  Zero

  - The bottom of the ladder represents the worst possible life for you.

---

```{r}
library(tidyverse)
library(ggplot2)
library(countrycode)

```


## What Libraries did you use?

- library(tidyverse) #Use to tidy and filter the data with dplyr
- library(ggplot2) #Use to draw plots
- library(countrycode) #Use to categorize countries by continent

---


## Look at the data 


```{r}
glimpse(data1)

```

---

## Data Key Terms and Categories

- GDP per capita- (variable name gdp) in purchasing power parity.

- Social support- "If you were in trouble, do you have relatives or friends you can count on to help you?"

- Freedom to make life choices- Are you satisfied or dissatisfied with your freedom to choose what
you do with your life?"

---

## Missing Data

  - Compared to the large dataset there was only a handful of missing values that were easily omitted.
  
```{r}  

sapply(data1, function(x) sum(is.na(x)))
```

  - Due to limited time and multiple columns, I will not explore the columns of generosity, positive affect, and negative affect.


```{r}
top_2009 = data1 %>%  select(Country.Name, Year, Life.Ladder) %>% 
  filter(Year == 2009) %>% slice_max(Life.Ladder, n=10)

lowest_2019 = data1 %>%   select(Country.Name, Year, Life.Ladder) %>% 
  filter(Year == 2019) %>% slice_min(Life.Ladder, n=10)


```

---

## Filter the data to year 2019 and find the max

```{r}
top_2019 = data1 %>%   select(Country.Name, Year, Life.Ladder) %>% 
  filter(Year == 2019) %>% slice_max(Life.Ladder, n=10)

```


```{r}
top_2019_plot= ggplot(top_2019, aes(reorder(Country.Name, Life.Ladder),Life.Ladder,fill=Country.Name)) + geom_col()+ theme(axis.text.x = element_text(angle=45))+
  ggtitle("Top ten in 2019")+xlab("Country")
top_2019_plot
```

---

## Now find the minimum in 2019

```{r }

lowest_2019_plot= ggplot(lowest_2019, aes(reorder(Country.Name, Life.Ladder),Life.Ladder,fill=Country.Name)) + geom_col()+ theme(axis.text.x = element_text(angle=45))+ggtitle("Lowest Life.Ladder scores in 2019")+xlab("Country")
lowest_2019_plot
 
```

---


## The countries with the highest happiness scores most frequently

```{r}


top = data1 %>%  select(Country.Name, Year, Life.Ladder) %>% 
  group_by(Year) %>% 
  slice_max(Life.Ladder, n=10)

top_all_years=top %>% group_by(Country.Name) %>% 
  count(Country.Name) %>% arrange(desc(n)) %>% filter(n > 8)

most_times_top_10_plot= ggplot(top_all_years, 
  aes(reorder(Country.Name, n),n,fill= n))  + geom_col()+
  theme(axis.text.x = element_text(angle=45))+
  ggtitle("Countries that made the Top Ten Most Frequently")+
  xlab("Country")+ ylab('Number of Times')
most_times_top_10_plot

```

---

```{r}
low = data1 %>%  select(Country.Name, Year, Life.Ladder) %>% 
  #group_by(Country.Name) %>% 
  slice_min(Life.Ladder, n=10)

```

## This dataframe is looking at how many times the countries scored the lowest on Life. Ladder.  
  - It uses the count function.
  - Then, on the next slide, we will take a look at the plot.
```{r}
low_all_years= low %>% group_by(Country.Name) %>% 
  count(Country.Name) %>% arrange(desc(n)) 
low_all_years
```

---


## The countries with the lowest happiness scores most frequently

```{r}
most_times_low_10_plot= ggplot(low_all_years, 
  aes(reorder(Country.Name, n),n,fill= n))  + geom_col()+
  theme(axis.text.x = element_text(angle=45))+
  ggtitle("Countries that had the Lowest Happiness Score Most Frequently")+
  xlab("Country")+ ylab('Number of Times')
most_times_low_10_plot


```

---

- Let's see if there is a relationship between having confidence in the national government and overall well-being.

```{r}
overall <- ggplot(data1, aes(Confidence.in.national.government, Life.Ladder, color = Life.Ladder, size = Life.Ladder)) + geom_point(alpha=1/3,size = 1)+ geom_smooth()
overall
```

- This chart is overplotted and it is difficult to see a 
clear relationship.  
- Let's filter the data to make it more clear.

---

# This data is filtered to year 2021 and shows the countries with the highest confidence in the national government.

```{r}

max_govt=data1 %>% filter(Year == 2021) %>% 
  select(Country.Name,
         Life.Ladder,Confidence.in.national.government,Year) %>%  slice_max(Confidence.in.national.government, n =12) 
max_govt


```



---

```{r}

low_govt = data1 %>% filter(Year == 2021) %>% 
  select(Country.Name,
         Life.Ladder,Confidence.in.national.government,Year) %>% slice_min(Confidence.in.national.government, n =15) 

```


## Here is a plot showing the relationship between high confidence in the Govt and happiness score
    - There doesn't appear to be a relationship

```{r}

high <- ggplot(max_govt, aes(Confidence.in.national.government, Life.Ladder, color = Life.Ladder, size = Life.Ladder)) + geom_point()+geom_smooth()
high
```

---


## Here is a plot showing the relationship between low confidence in the Govt and happiness score.
    - The happiness score appears to be increasing as confidence increases.

```{r}
low <- ggplot(low_govt, aes(Confidence.in.national.government, Life.Ladder, color = Life.Ladder, size = Life.Ladder)) + geom_point()+ geom_smooth()
low
```

---


## Bolivia & USA

- I have heard that there is a high perception of corruption in Bolivia.  I wanted to explore the relationship with a plot.


```{r}

US_Bolivia = data1 %>% filter(Country.Name == 'Bolivia'|Country.Name == 'United States')
Bolivia_corruption_plot = ggplot(US_Bolivia,aes(Perceptions.of.corruption, Life.Ladder,color=Country.Name, size= 3))+geom_point()
Bolivia_corruption_plot
```

---


## Does healthy Life Expectancy have a relationship with Well-Being?

- Let's explore the difference between Bolivia and the USA

```{r}
Bolivia_lifeexp_plot = ggplot(US_Bolivia,aes(Healthy.life.expectancy.at.birth,
 Life.Ladder,color=Country.Name, size= 3))+geom_point(alpha= 2/3)
Bolivia_lifeexp_plot

```

---

# The original dataset did not have a category for continent, however RStudio has a library called countrycode which can be used to categorize by continent.  
- It uses the five continent model:


```{r Continent Code, echo=TRUE}

##Five Continent Model

data1$Continent <- countrycode(sourcevar = data1$Country.Name,
                             origin = "country.name",
                             destination = "continent")

data1$Continent[data1$Country.Name == "Kosovo"] = "Europe"
 
by_continent = data1 %>% select(Country.Name,Continent, Life.Ladder) %>% arrange(Continent) %>% group_by(Continent)

```

---



## Continent DataFrame

- a dataframe showing the Life.Ladder for each continent
    - average 
    - max 
    - min 

```{r}
by_continent = data1 %>% select(Country.Name,Continent, Life.Ladder) %>% arrange(Continent) %>% group_by(Continent) %>% summarize(avg = mean(Life.Ladder),max=max(Life.Ladder),min = min(Life.Ladder))
by_continent
```

---


## Because of the closeness in Oceania's numbers, let's take a closer look at Oceania to see which countries it represents in this dataset.

```{r Oceania, echo=TRUE}
Oceania =data1 %>% filter(Continent == 'Oceania') %>% count(Country.Name)
Oceania
```
- Now we know it only represents two countries.

---



## Average happiness score by Continent

```{r}

continent_plot= ggplot(by_continent, 
  aes(Continent, avg,fill=avg))  + geom_col()+
  theme(axis.text.x = element_text(angle=45))+
  ggtitle("Comparison of Average Life.Ladder by Continent")+
  xlab("Continent")+ ylab('Happiness (Life.Ladder)')
continent_plot
  
```

---



## Comparison of Max Life.Ladder score by Continent

```{r}
continent_plot= ggplot(by_continent, 
  aes(Continent, max,fill=max))  + geom_col()+
  theme(axis.text.x = element_text(angle=45))+
  ggtitle("Comparison of Max Life.Ladder by Continent")+
  xlab("Continent")+ ylab('Happiness (Life.Ladder)')
continent_plot
```

---


## Plot of Min Life.Ladder scores by Continent

```{r}
continent_plot= ggplot(by_continent, 
  aes(Continent, min,fill=min))  + geom_col()+
  theme(axis.text.x = element_text(angle=45))+
  ggtitle("Comparison of Min Life.Ladder by Continent")+
  xlab("Continent")+ ylab('Happiness (Life.Ladder)')
continent_plot
```

- Note that Oceania has only a small difference between the max and min. 
- Oceania also has the fewest participants of all the continents.(Only 2 countries)

---

```{r}
gdp_ = data1 %>%  select(Country.Name, Year, Life.Ladder,Log.GDP.per.capita) %>% 
  group_by(Year) %>% 
  arrange(desc(Log.GDP.per.capita)) %>% 
  filter(Year==2019|Country.Name=='United States')

```

## Compare the United States GDP to Life.Ladder


```{r message=FALSE, warning=FALSE}
gdp_plot= ggplot(gdp_, aes(Life.Ladder, Log.GDP.per.capita, fill=Log.GDP.per.capita))  + geom_point() + geom_smooth()
  theme(axis.text.x = element_text(angle=15))+
  ggtitle("United States: Happiness score versus GDP")+
  xlab("Happiness (Life.Ladder)") + ylab('GDP')
gdp_plot


```

- According to the data, Life.Ladder generally increases with GDP.


---

# This dataframe shows the relationship between social support and the feeling of well-being.

```{r}
Social_Support = data1 %>% na.omit() %>% select(Country.Name,Continent, Life.Ladder, Social.support) %>% arrange(Continent) %>% group_by(Continent) %>% summarize(avg_social_support = mean(Social.support),median_life_ladder=median(Life.Ladder)) %>% arrange(avg_social_support)
Social_Support
```

---


## A box plot of each Continent showing social support on the x-axis and life ladder on the y-axis

```{r}
data1 %>% 
  ggplot(mapping = aes(x = Social.support, y = Life.Ladder)) +
  geom_boxplot()+facet_wrap(~Continent)


```